Transport map accelerated-PAIS, and application to inverse problems arising from multiscale stochastic reaction networks

### 2018-01-19 TPAIS_2018.tex
Added huge amounts + edits
Paper now a monster.

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)

In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

@@

\section{Multiscale Methods for Stochastic Chemical Reaction
  Networks}\label{sec:multi}
In this Section, we discuss some recent advances in multiscale methods
for stochastic reaction networks. Inverse problems arising in this
area often lead to highly correlated and complex posterior
distributions, which traditional MCMC methods can struggle to sample
from. We will then go on to solve some inverse problems related to
this in Section \ref{sec:num}.

\section{Numerical Examples}\label{sec:num}

>>>>

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)
In this Section, we describe the construction of transport maps which
allow for the simplification of complex posterior distributions in
order to allow for improved sampling, in particular for methods based
on importance sampling.
In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

@@

\section{Multiscale Methods for Stochastic Chemical Reaction
  Networks}\label{sec:multi}
In this Section, we discuss some recent advances in multiscale methods
for stochastic reaction networks. Inverse problems arising in this
area often lead to highly correlated and complex posterior
distributions, which traditional MCMC methods can struggle to sample
from. We will then go on to solve some inverse problems related to
this in Section \ref{sec:num}, using the transport map versions of the
PAIS algorithm, as described in Section \ref{sec:TPAIS}.

We consider chemical reaction networks of $N_s$ chemical species $\{S_j\}_{j=1}^{N_s}$,
with population numbers given by $X(t) = [X_1(t), X_2(t), \ldots, X_{N_s}(t)]^\top \in
\mathbb{N}_0^{N_s}$ reacting in a small reactor, through $N_r$ different reaction
channels. When population numbers of one or more of the chemical
species in the reactor is small, as is the case with chemical
reactions occurring within living cells, the sporadic and discrete way
in which reactions occur can not be well modelled by deterministic
continuous models, such as ODEs. In such a situation, we may wish to
model the dynamics through a discrete stochastic model, such as a
continuous time Markov chain.

For each reaction $R_j$, for $j = 1,2,\ldots N_r$, there is a
propensity or hazard function $\alpha_j(X(t)$ which indicates how
likely that reaction is to fire, defined by
\[\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
    the time interval  } \quad s \in [t, t+ dt] ).\] If a system
satsifies what is called \emph{mass action kinetics}, then the form of
the function $\alpha_j$ is determined, up to a rate constant, by the
reactants involved in that reaction:
\begin{equation}\label{eq:MAK}
\alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
\end{equation}
where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
$\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
$k_j$ are rate constants.

Following each reaction $R_j$ there is an instantaneous change in the
current state, as the reactants of the reaction are consumed, and the
products produced. This is modelled by a change in the state vector
$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stochiometric vector for
reaction $R_j$.

The model can be represented as the following expression involving
$N_r$ different unit rate Poisson
processes****CITE ANDERSON ET AL**** $Y_j$, given by:
\begin{equation}\label{eq:RTC}
X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
\end{equation}

The master equation for this model is only solvable in certain
situations, for example monomolecular networks****CITE****, or for
steady-state solutions certain deficiency
zero networks****CITE*****. Trajectories for this system can be
sampled exactly, for instance using the Gillespie SSA***CITE***, or
its variants\cite{gillespie2007stochastic}****CITE****. However, if the system is stiff, i.e. there
are some reactions which are firing many times on a timescale for
which others are unlikely to fire at all, then trajectories can become
prohibitively expensive to simulate, since these methods simulate
every single reaction event with the same cost. In such a system, one
might employ multiscale methods in order to approximate trajectories
at a lower cost than the exact algorithms.

One common approach is to induce the quasi-steady-state approximation
QSSA. This approximation makes the assumption that fast reactions
enter quasi-equillibrium on a timescale which is negligible with
respect to the timescale on which the slow dynamics in the system are
evolving. This amounts to taking the assymptotic limit that the
propensities of the ``slow'' reaction channels are equal to zero. This
allows us to sample approximate trajectories of the slow reactions
without the need to compute many fast reaction events.

This approach can work well when there is a large timescale gap
between the fast and slow reactions in a system, but where the
timescale gap is less pronounced, it can lead to significant
errors***cite cotter erban errors paper***. Another approach is the
constrained multiscale method (CMA)***CITE***, based in part on the
equation-free approach to multiscale computations***CITE
YANNIS****. This approach also assumes quasi-equillibrium in the
system, but takes into account the effect that the slow reactions can
have on the invariant distribution of the fast variables in the
system. For a more detailed description of this method, please refer
to the literature***CITATIONS***.

Multiscale methods are not only of use when forward evaluations of the
dynamics are costly, but can also be of use where we attempt to solve
an inverse problem where the fast variables are unobservable. One
approach in this situation would be to integrate over all possible
trajectories of the fast variables, but this would almost always be
prohibitively expensive. Another approach would be to use an
appropriate multiscale approximation, so that the effective dynamics
of the slow variables can be approximate without the need to consider
the rapid fluctuations of the fast variables in the system.

\subsection{Likelihoods arising from stochastic reaction networks}
Suppose that we are able to exactly observe the number of molecules of
each chemical species in a system which satisfies mass action
kinetics, and which can be well approximated by
\eqref{eq:RTC}. Suppose that we wish to be able to infer the value of
the rate constants of each reaction from these observations. Even with no observational noise, since the dynamics
of the system are stochastic, this still leads to a Bayesian inverse
problem where we can only infer a joint probability distributions on the
reaction parameters.

Suppose that we are in state $X(t) = X_0$. There are two independent
univariate random variables which decide when and what the next
reaction in the system is. There is the $j$th waiting time $\tau_j$ to the
next reaction, which is given by an exponential random variable
\[\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\]
where $\alpha_0(X(t)) =\sum_{i}^{N_r} \alpha_i(X(t))$ is the total
propensity in the system. The second is a multinomial random variable $r_j$
which dictates which reaction has occurred during the $j$th reaction,
which takes the value
$r \in \{1,2,\ldots,N_r\}$ with probability
\[\mathbb{P}(r = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
As such, in order to compute the likelihood of a particular trajectory
given a possible realisation of the reaction parameters, it is
sufficient to have access to the total time spent in each state, and
the frequency of each reaction which led to leaving each state.

From this formulation, we see that the random variables $(\tau_j, r_j)$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are Markovian. This conditional independence means that we can group events together by what state the system was in when the event happened. We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
\[
	T(\mathbf{Y}) = \sum\limits_{j=1}^M \tau_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
\]
This random variable, $T(\mathbf{Y})$, is a sum of exponentially
distributed random variable, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
\begin{equation}\label{eqn:chem_time_dist}
	T(\mathbf{Y}) \sim \text{Gamma}\left(\alpha=K(\mathbf{Y}),~\beta = \alpha_0(\mathbf{Y}; \mathbf{k})\right),
\end{equation}
where $K(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y})$.

Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_{N_r}(\mathbf{Y})]^\top$ where
\[
	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r_j = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
\]
Here all the random variables $r_j$ follow the same multinomial distribution, and so
\begin{equation}\label{eqn:chem_react_dist}
	\mathbf{r}(\mathbf{Y}) \sim \text{Multinomial}(K(\mathbf{Y}),~\mathbf{p}(\mathbf{Y})). 
\end{equation}
The random variables defined in Equations~\eqref{eqn:chem_time_dist} and \eqref{eqn:chem_react_dist} are sufficient statistics for the posterior distribution $\pi(\mathbf{k}|D)$. With these definitions we define two new structures
\[
	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
\]
where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
\[
	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
\]

To construct the posterior distribution for the reaction rates, $\mathbf{k}$, in the chemical system, we formulate the likelihood using the sufficient statistics derived in the previous section. Due to the non-negativity of these reaction rates, we assign a Gamma prior distribution to each rate. Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
\[
	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
\]
where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities
$\alpha_i$ and probabilities $p_i$ depend on the reaction rates
$\mathbf{k} = [k_1, k_2, \ldots k_{N_r}]^\top$ through the concept of
mass action kinetics given in \eqref{eq:MAK}.

Our choice of Gamma prior distributions with hyper-parameters $(a_i, b_i)$ results in the posterior distribution of the form
\begin{align}
	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
%		& \qquad\qquad\qquad
				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
			\Bigg]  \notag\\
		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eqn:chem_posterior}
\end{align}

\subsection{Approximation of likelihoods in multiscale chemical
  networks}
Suppose now that we are only observing the slower variables in a
multiscale chemical system of this type. Suppose that $n_r < N_r$ is the number of slow
reactions in the system, and the slow reactions are given by $\{R_{s_1},
R_{s_2}, \ldots R_{s_{n_r}} \} \subset \{R_1, R_2, \ldots , R_{N_r}
\}$. The propensities of the slow
reactions $\alpha_{s_i}$ may depend on the value of the fast variables which is
unknown. However, the invariant distribution of the fast variables
conditioned on the slow variables in the system can be approximated
using a multiscale method, such as the QSSA or the CMA, as described
briefly in Section \ref{sec:multi}. Once the approximations have been
made, we arrive at approximate \emph{effective} propensities $\bar{\alpha}_{s_i}$, which
have been averaged over the computed invariant distributions of the
fast variables. Then the approximate effective dynamics on the slow variable $S(t)$ are given by 
\begin{equation}\label{eq:RTCS}
S(t) = S(0) + \sum_{j=1}^{n_r} \nu_{s_j} \int_0^t \bar{\alpha}_{s_j}(S(s)) ds.
\end{equation}

In turn, the approximate posterior distribution on the reaction
parameters $\mathbf{k}$ is then given by
\begin{align}
	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
%		& \qquad\qquad\qquad
				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
			\Bigg]  \notag\\
		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eq:Apos}
\end{align}

\section{Numerical Examples}\label{sec:num}
In this section we look at two examples of chemical systems to
demonstrate the effectiveness of the Bayesian approach we have
proposed, along with the transport PAIS algorithms for sampling the
challenging posterior distribution on the reaction parameters that
arise.

% The first example will only contain at most mono-molecular reactions,
% i.e. reactions which convert one species into another species at a
% particular rate. As shown in~\cite{jahnke2007solving} these systems
% have a Poisson invariant distributions on the populations, given by
% \begin{equation}\label{eqn:mono_stat_dist}
% 	\mathbb{P}(X_1=x_1, \ldots, X_n=x_n) = \left(\prod_{i=1}^n \! \frac{\bar{c}_i^{x_i}}{x_i!}\right)\exp\left(-\sum_i^n \! \bar{c}_i\right),
% \end{equation}
% where $x_i \in \mathbb{N}$ and $\bar{c}_i$ are the complex balances or the concentration of chemical $X_i$ according to the solution of a steady state approximation of the system. We make use of this distribution to formulate posteriors for the chemical reaction rates using the constrained multiscale approach, as in~\cite{cotter2016constrained}.

%%%%%

\subsection{A Multiscale Chemical System}\label{sec:chem_multiscale}

First we consider a simple multiscale example involving two chemical
species $S_1$ and $S_2$ involved in only zeroth and first order reactions:
\begin{equation}\label{eqn:full_system}
	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
\end{equation}
Each arrow represents a reaction from a reactant to a product, with
some rate constant $k_i$, and where mass action kinetics is
assumed. The parameters $k_i$ are non-negative, and $\mathbf{k} =
[k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the
population count of
species $S_i$ by $X_i \in \mathbb{N}_0$. We assume that we are in a
parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
chemical species are involved in fast reactions. However, the quantity
$S = X_1 + X_2$, is conserved by both of the fast reactions,
and as such, this is the slowly changing quantity in this system.

We assume that we are only able to observe
the slow variable $S(t)$. 
The effective dynamics of $S$ can be approximated by a system of only
two reactions:
\begin{equation}\label{eqn:QSSA_system}
	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset.
\end{equation}
The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
application of a multiscale approximation, as detailed in Section
\ref{sec:multi}, and in more detail in ****CITE****. Under the QEA,
this is given by
\[
	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
\]
Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
\begin{equation}\label{eqn:chem_CMA_rate}
	\bar{\alpha}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
\end{equation}

Our observations are uninformative about the reaction rates $k_2, k_3$, and $k_4$, as there
is a sub-manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
problem. This type of problem is notoriously difficult to
sample from using standard MH algorithms, as the algorithms quickly
gravitate towards the manifold $\mathcal{M}$ but then
exploration around $\mathcal{M}$ is slow.

We aim to recover three different posterior distributions. In the
first, which is of the form \eqref{eqn:chem_posterior} we assume that we can fully and exactly observe the whole state
vector for the whole of the observation time window. In the second and
third, the observations are restricted to observations of the slow
variable $S = X_1 + X_2$. In these two examples, we approximate the
posterior using \eqref{eq:Apos} with QEA and CMA approximations of the
effective dynamics respectively.

In all three cases, we find the posterior distribution for $\mathbf{k} \in \mathcal{X}$. These four parameters are assigned Gamma prior distributions,
\[
	k_i \sim \text{Gamma}(\cdot; \alpha_i, \beta_i), \quad \text{for} \quad i = 1, \dots, 4.
\]
The hyper-parameters corresponding to each of these prior distributions are given in Table~\ref{tab:multiscale_priors}. These priors are the same for each of the three posterior distributions.

\begin{table}
\centering
\begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
	\hline
	parameter/dimension & 1 & 2 & 3 & 4 \\ \hline
	$\alpha_i$ & 150 & 5 & 5 & 3 \\
	$\beta_i$ & 15/9 & 5/12 & 5/12 & 1 \\ \hline
\end{tabular}
\caption{Hyper-parameters in the prior distributions for the multiscale problem described in Section~\ref{sec:chem_multiscale}.}
\label{tab:multiscale_priors}
\end{table}

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
\caption{Posterior for the constrained multiscale problem outlined in Section~\ref{sec:chem_multiscale}.}
\label{fig:chem_CMA_posterior}
\end{figure}

Figure~\ref{fig:chem_CMA_posterior} shows how this posterior looks
when we use the CMA to model the effective degradation propensity $\bar{\alpha}_4$.

We consider several different algorithms for sampling from these
distributions. First we implement both the PAIS and MH algorithms with
a Gaussian proposal distribution. In the case of the PAIS algorithm,
this is a Gaussian mixture proposal distribution, and in the MH
algorithm this is equivalent to a vanilla random walk. The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm. We also compare the PAIS and MH algorithms when using a transport map proposal distribution. This proposal method was discussed in detail in Chapter~\ref{sec:TPAIS}.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\pgfmathsetseed{1}

		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};

		\draw (A) \irregularcircle{1cm}{2.5mm};
		\draw (B) \irregularcircle{1cm}{0.5mm};

		\draw [->] (A) -- (B);

		\node at (2.5,1) {$T\colon\mathcal{X}\rightarrow R$};

		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
	\end{tikzpicture}
	\caption{Couplings between parameter space and reference when using the transport map proposal method to propose moves on parameter space.}
	\label{fig:chem_coupling}
\end{figure}

Figure~\ref{fig:chem_coupling} shows how we define a bijective map,
$\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
$R$. When using the transport map proposal distribution, we prefix the proposal method with a T, e.g. MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW. In practice we cannot ensure that the approximate map $\tilde{T}$
is uniquely invertible over the whole of $R$ and so $\tilde{T}$ is not
truly bijective. This leads to problems for our strictly positive
state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
$R$ do not necessarily map back onto $\mathcal{X}$. This motivates the
use of an intermediate space.

We also consider how these algorithms perform when they are applied to
the log of the reaction rates. We choose to use this log
transformation since it converts our strictly positive parameter
space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
\mathbb{R}^d$. This allows us to define $T$ between two subsets of
$\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely
invertible in some region of $R$, all possibilities are valid
proposals. It also reduces the work required of the map, which since
we use only a finite dimensional approximation, can improve
performance and stability. As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g. MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\pgfmathsetseed{1}

		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};

		\draw (A) \irregularcircle{1cm}{2.5mm};
		\draw (B) \irregularcircle{1cm}{1.1mm};
		\draw (C) \irregularcircle{1cm}{0.5mm};

		\draw [->] (A) -- (B);
		\draw [->] (B) -- (C);

		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};

		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
	\end{tikzpicture}
	\caption{Couplings between parameter space, an intermediate space and reference space when using the transport map proposal method to propose moves on a log space.}
	\label{fig:chem_log_coupling}
\end{figure}

Figure~\ref{fig:chem_log_coupling} displays the composition of maps when we include this intermediate space $\mathcal{Y}$. Every proposal in $R$ results in a valid proposal on $\mathcal{X}$. The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$. The weight is now
\[
	w_i(\theta) = \frac{\pi(\theta|\mathbf{R},\mathbf{T})}{\chi(\tilde{T}\circ\log(\theta)|\tilde{T}\circ\log(\theta^{(i-1)}))|J_{\tilde{T}\circ\log}(\theta)|},
\]
where $\theta$ is a proposed new state, $\theta^{(i-1)}$ is the ensemble of states from the previous iteration, and $J_{\tilde{T}\circ\log}(\theta)$ is the Jacobian of the composition of the two maps. This Jacobian is straightforward to calculate,
\[
	|J_{\tilde{T}\circ\log}(\theta')| = |J_{\tilde{T}}(\log(\theta'))||J_{\log}(\theta')|,
\]
where the first determinant is as we saw in the previous chapter, and the second is
\[
	|J_{\log}(\theta')| = \prod\limits_{i=1}^d \frac{1}{\theta'_i}.
\]

% choices made in transport map
For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
\[
	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
\]
and
\[
	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
\]

% choice of resampler?
We will use the AMR***CITE*** resampler with an ensemble size of
$M=500$. This increase in ensemble size in comparison with previous
example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.

% analysis method
To measure the convergence of the sampling methods in this section, we
will compare the convergence of the mean of each parameter. We
approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from
the MH-RW algorithm. We do this for each of the eight algorithms we
have so far discussed. Convergence is shown only for the CMA
approximation of the posterior \eqref{eq:Apos} with effective rate for
the degradation of $S$ given by \eqref{eqn:chem_CMA_rate}, but we
expect very similar results for the other posterior distributions discussed.

\begin{table}[!h]
\centering
\begin{tabular}{rrrrrr}
\toprule
	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{PAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
	& & $\delta_\%$ & & $\delta_{\text{ESS}}$ & ESS \\ \midrule
	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
	TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
\bottomrule
\end{tabular}
\caption{Optimal scaling parameters for the MH and PAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}. MH parameters optimised by acceptance rate, and PAIS parameters optimised using effective sample size.}
\label{tab:chem_multiscale_scaling}
\end{table}

The optimal scaling parameters for the proposal distributions are given in Table~\ref{tab:chem_multiscale_scaling}. We note that for MH-TRW and PAIS-TRW the scaling parameter is near to 1, particularly for MH-RW, this is what we should expect since the proposal is made on a reference space which should be near to $\mathcal{N}_d(0, \text{I})$. The same should be true for the MH-logTRW and PAIS-logTRW algorithms since these have the same target reference space, however the optimal scaling parameters here are much smaller. We see that the ESS is higher for the algorithms which sample on $\mathcal{Y}$, and we expect that convergence will be fastest for the PAIS-logTRW algorithm.

\begin{figure}[!htb]
\centering
\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
\caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
\label{fig:chem_multiscale_L2}
\end{figure}

Convergence of the eight algorithms for this example is shown in Figure~\ref{fig:chem_multiscale_L2}. We first note the poor performance of the MH based algorithms, each of them taking roughly 10,000 samples to begin converging. Only the MH-TRW is at all competitive with the PAIS algorithms. During the simulation interval, the MH-TRW algorithm has not settled down to the expected $\mathcal{O}(1/\sqrt{N})$ rate which means that the estimate is still biased by the long burn-in time. As we have seen in previous examples, the burn-in time for the PAIS algorithm is negligible.

The PAIS variants with RW and TRW proposals perform similarly on both sample spaces. When sampling on $\mathcal{X}$ the transport map is not quite as efficient, largely due to the difficulties discussed in the previous section i.e. many proposals are made which do result in negative reaction rates. Sampling on $\mathcal{Y}$ leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable. This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$. Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the PAIS-logTRW algorithm.

\begin{figure}[!htb]
\centering
\subfigure[Reference space mapped to by $\hat{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
\subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
\caption{Reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions. Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
\label{fig:chem_reference_spaces}
\end{figure}

We now look at the proposal distributions of the transport map accelerated algorithms. In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; in (a) mapping the posterior through the map $\hat{T}$, and in (b) by mapping the posterior through $T\circ\log$. For the most part, each of these marginal distributions can be recognised as a Gaussian. However, with the exception of $\mathbb{P}(r_2,r_3)$, we would not consider them to be close to a standardised $\mathcal{N}(0, \text{I})$ distribution. Before thinking that the transport map has not helped us to find a `nicer' space on which to propose new values we should consider that the dimensions are now (1) largely uncorrelated, and (2) the variances in each dimension are much more similar than they are in Figure~\ref{fig:chem_CMA_posterior}.

Particularly in Figure~\ref{fig:chem_reference_spaces} (b) we see that $\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than $\text{var}(r_2)$ and $\text{var}(r_3)$. To combat this we have a number of choices, we might wish to use two different scaling parameters to match these scales, which would require knowledge of the reference space before beginning sampling. We could alternatively use a proposal distribution such as in the LPAIS algorithm to adaptively learn this at each iteration. Another option is to increase the total order of our index set. For these numerics we have chosen $p=3$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.

\subsubsection{Comparison of the Constrained and QEA approaches}

The convergence analysis has been performed for the constrained approach to this multiscale system. We now look at the differences between the constrained and QEA posterior distributions. Recall that the approaches differed only in the form of the effective degradation rate $\hat{k}_4$,
\[
	\hat{k}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{k}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
\]
This difference in the denominator causes a shift in the parameters as can be seen in Figure~\ref{fig:chem_diff}. The figure shows the difference in posteriors,
\begin{equation}\label{eqn:chem_diff}
	\text{diff}(\mu^{\text{CMA}}, \mu^{\text{QEA}}) = \pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
\end{equation}

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
\caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
\label{fig:chem_diff}
\end{figure}

Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive. These positive differences tell us that the constrained approach claims to be more informative about the values of these first three parameters. However the marginal for $k_4$ varies by a significant amount and is more negative. This means that the mean estimates for $k_4$ differ by a large amount, the posterior for the QEA is most peaked.

% compare solution with full data.
We now consider how we should interpret the information given by these models. The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ independently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$. Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$. To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most accurate and informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model, with all reactions observed.
